# Introduction

Welcome to the training manual on using R, RStudio, Git, and GitHub for Household Consumption and Expenditure surveys. This manual is designed for absolute beginners and aims to provide a comprehensive introduction to these powerful technologies.

First, we will cover R, which is a powerful and versatile programming language that is widely used for data analysis, statistical modeling, and data visualization. 

-   It is an open-source software that can be freely downloaded and used by anyone. R is widely used in academia, industry, and government, and is becoming increasingly popular among data scientists and analysts. 
-   It is a great tool for those who have been using other statistics tools like Excel, SAS, SPSS and want to take their data analysis skills to the next level.

This training will provide an introduction to the basics of R and will give you the skills you need to start working with data in R..

Next, we will introduce RStudio, which is a popular integrated development environment (IDE) for R. 

-   RStudio provides a user-friendly interface for working with R and makes it easy to work with R packages, which are collections of pre-written R code that can be used to perform specific tasks. 
-   With RStudio, you will be able to write, test, and debug your R code, and easily share your work with others.

<!-- Git and GitHub will be covered next. Git is a version control system that allows developers to track changes in their code and collaborate on projects with others. 

-   It is widely used by software developers to manage their source code and collaborate with others on projects. 
-   Git allows developers to easily create new versions of their code, revert to previous versions, and merge changes from multiple contributors. 

GitHub, which is a web-based platform that allows developers to host, share and collaborate on their Git repositories, provides a central place for storing and managing code, as well as tools for working with others on the same codebase. 

-   With Git and GitHub, you will be able to easily manage your code, collaborate with others, and share your work with the world. -->

This manual will provide step-by-step instructions for installing and setting up R and RStudio on your computer. We will also go over basic concepts and commands for working with each technology, as well as provide examples of how to use them in different contexts. With this manual, you will have the skills and knowledge you need to start using these powerful technologies to better understand your data and collaborate with others on your project.

## Software requirements

### Downloading and Installing R
To download R, you can visit the official R website at [https://cran.r-project.org/](https://cran.r-project.org/). On the website, you will see links to download the latest version of R for Windows, Mac, and Linux. Once you have downloaded the installer for your operating system, you can run the installer and follow the prompts to install R on your computer.

#### Updating R
If you already have R installed on your computer, you can check if you have the latest version by opening RStudio and clicking on the “Help” menu. If you see a message that says “A new version of R is available”, you can click on the “Install Update” button to update R to the latest version. If you do not see this message, you can check the version of R by typing the following command in the R console:

```{r eval=TRUE}
R.version.string
```

If the version number is less than the one shown above you need to update R to the latest version. You can do this from within RStudio by following the steps below:

```{r}
#1. Install the package "installr
install.packages("installr")

# 2. load the package "installr" from the library: 
library(installr)

# 3. Update R: 
updateR()
```

### Downloading and Installing RStudio
To download RStudio, you can visit the official RStudio website at [https://rstudio.com/products/rstudio/download/](https://rstudio.com/products/rstudio/download/). On the website, you will see links to download the latest version of RStudio for Windows, Mac, and Linux. Once you have downloaded the installer for your operating system, you can run the installer and follow the prompts to install RStudio on your computer.
<!-- 
### Downloading and Installing Git
To download Git, you can visit the official Git website at[https://git-scm.com/downloads](https://git-scm.com/downloads). On the website, you will see links to download the latest version of Git for Windows, Mac, and Linux. Once you have downloaded the installer for your operating system, you can run the installer and follow the prompts to install Git on your computer. -->

Please note that these are general instructions, and depending on your system setup and security settings, some steps might be slightly different. Also, you will need to make sure that you have administrative access or permission to install the software on your computer.

You can also refer to the software website instruction or online tutorials that are specific to your operating system and setup.
