


# Basic Statistics: Jura

:::{.callout-note}
## Exercise: Exploratory data analysis


Run the blocks below and:
i)  Discuss the summary statistics and distributions
i)  Carry out the transformations and discuss the results
:::

### Import custom functions
```{r eval=TRUE}
source("data/custom_functions/CEPHaStat_3.R")
source("data/custom_functions/boxcox.R")
```

### Import Jura data
```{r eval=TRUE}
data.df<-read.csv("data/Jura.csv",header=T,stringsAsFactors=T)
```

###  Jura Pb Analyses

#### Distributions and Summary statistics

```{r eval=TRUE}
summa(data.df$Pb)
summaplot(data.df$Pb,"Pb /ppm")
```

:::{.callout-tip}
## Remember to base transformations on residuals, after fitting the model!
:::
    
```{r eval=TRUE}
mod<-lm(Pb~Rock,data=data.df)
summa(mod$residuals)
summaplot(mod$residuals,"Residual /ppm Pb")
```


#### Comparing the log and BoxCox transformations

:::{.callout-tip}
## Remember to base transformations on residuals, after fitting the model!
:::

#### log transformation

```{r eval=TRUE}
mod2<-lm(log(Pb)~Rock,data=data.df)
summa(mod2$residuals)
summaplot(mod2$residuals,"Residual /Bcox ppm Pb")
anova(mod2)
```

#### BoxCox transformation

```{r eval=TRUE}
mod3<-lm(BoxCox(Pb)~Rock,data=data.df)
summa(mod3$residuals)
summaplot(mod3$residuals,"Residual /Bcox ppm Pb")

anova(mod3)
```


#### Co analyses

```{r eval=TRUE}
summa(data.df$Co)
summaplot(data.df$Co,"Co /ppm")

mod<-lm(Co~Rock,data=data.df)

summa(mod$residuals)
summaplot(mod$residuals,"Residual /ppm Co")
```

:::{.callout-note}
# What do you conclude from the Co analysis?
:::


